---
title: "Stat 201 - Lecture 10"
author: "Prof. Heggeseth"
date: "October 15, 2015"
output: slidy_presentation
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(dev='png',fig.path='Figs/',echo=TRUE,message=FALSE, warning=FALSE,message=FALSE,tidy=TRUE,fig.width = 6,fig.height=6)
```

----------

#Babies

Let's think about babies again.

```{r, echo=FALSE}
require(mosaic)
par(mar=c(5,4,4,1)) #setting the margins of the plotting region
load('~/Dropbox/Shared Files/Data sets/BABIES1998.Rda')
babies1998dec <- babies1998[babies1998$dobmm == 12,] #[rows, columns]
```

```{r}
with(babies1998dec,hist(gestation,xlab='Gestation (weeks)',main='1998 Population',xlim=c(15,50)))
```

Let's imagine that we were going to gather data by random sampling 100 babies and find the mean gestational age.

We can simulate the sampling distribution by taking many random samples out of the population (all of size $n=100$).

```{r}
sampdist = do(1000)*with(sample(babies1998dec, 100), mean(gestation, na.rm = TRUE))
hist(sampdist[,1],xlab='Mean Gestation (weeks)',main='Sampling Distribution')
sd(sampdist[,1]) #measure of variability
mean(sampdist[,1]) #practically the population mean
(popmean = mean(babies1998dec$gestation,na.rm=TRUE))
```

----------

#Bootstrap Confidence Intervals

##One Sample from Population

Imagine that we only had 1 sample of 100 babies (not the whole population). 

```{r}
require(mosaic)
smp <- sample(babies1998dec, size = 100, replace = FALSE) #drawn from population without replacement
```

To get a sense of sampling variability, we would bootstrap this sample and make a 95% confidence interval. 

```{r}
boot1 = do(1000)*with(resample(smp),mean(gestation, na.rm = TRUE))
(ci1 = quantile(boot1[,1],c(0.025,0.975)))
ci1[2] - ci1[1] #width of the interval 
(ci1[2] - ci1[1])/sd(boot1[,1]) #about 4 SD's wide
sd(boot1[,1]) #An estimate of sampling variability
mean(boot1[,1]) #C entered at the original sample mean (not population mean)
```

Does this interval contain the "true mean" gestational age of the population?

```{r}
popmean > ci1[1] & popmean < ci1[2]
```

----------

#Bootstrap Confidence Intervals


##Second Sample from Population

Imagine that we only had a different sample of 100 babies (not the whole population). 

```{r}
smp2 <- sample(babies1998dec, size = 100, replace = FALSE) #drawn from population without replacement
```

To get a sense of sampling variability, we would bootstrap this sample and make a 95% confidence interval. 

```{r}
boot2 = do(1000)*with(resample(smp2),mean(gestation, na.rm = TRUE))
(ci2 = quantile(boot2[,1],c(0.025,0.975)))
ci2[2] - ci2[1] #width of the interval
(ci2[2] - ci2[1])/sd(boot2[,1]) #about 4 SD's wide
sd(boot2[,1]) #An estimate of sampling variability
mean(boot2[,1]) #Centered at the original sample mean
```

Does this interval contain the "true mean" gestational age of the population?

```{r}
popmean > ci2[1] & popmean < ci2[2]
```


----------

#Bootstrap Confidence Intervals


##Third, Fourth, etc. Sample from Population

Let's do this same process:

- Take one sample from the population without replacement
- Bootstrap that sample
- Calculate 95% confidence interval
- Check whether the true population mean is in that interval

```{r}
manysamples = function(){
  smp <- sample(babies1998dec, size = 100, replace = FALSE) #drawn from population without replacement
  boot = do(1000)*with(resample(smp),mean(gestation, na.rm = TRUE))
  (ci = quantile(boot[,1],c(0.025,0.975)))
  popmean > ci[1] & popmean < ci[2]
}


inCI = do(500)*manysamples() #takes a long time to run...

prop.table(table(inCI)) #Percent of time our interval "captures" the true mean
```

Thus, we can only be 95% confident that the "true mean" of the population is in our interval. 

- Some times it is, some times is it not (it depends on how "unusual" our sample is).
- This interval is giving us some sense of what the "true mean" is, not individual data points. 

Common Mistakes on Quiz:

- 95% of the days, she'll sell between 47 and 50 cups of soup.
- Each day she’ll sell between 47 and 49.5 soups per day.

