---
title: "Stat 201 - Lecture 20"
author: "Prof. Heggeseth"
date: "December 1, 2015"
output: slidy_presentation
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(dev='png',fig.path='Figs/',echo=TRUE,message=FALSE, warning=FALSE,message=FALSE,tidy=TRUE,fig.width = 6,fig.height=6)
```

--------------

#More Hypothesis Testing

- Experimental data with blocking
    - Paired data (Chp 23)
- Comparing many means
    - Adjust Type 1 error rate (pg 765)
    - Analysis of Variance (Chp 26)
- Compare more than 2 proportions
    - Tests of independence and homogeneity (Chp 25)
    - Goodness of Fit tests (Chp 25)
 
For all of these, we have $H_0$ and $H_A$, a test statistic, the sampling distribution for the test statistic, and a p-value. 

--------------

#An experiment

**Finger Tapping**

In 1994, Scott and Chen published research that compared the effects of caffein with those of theobromine (a similar chemical found in chocolate) and with those of a placebo. The experiment used 4 human subjects, and took place over several days. 

Each day, each subject swallowed a tablet containing one of caffeine, theobromine, or the placebo. 

Two hours later, they were timed while tapping a finger in a specified manner (they had practiced earlier). The resopnse is the number of taps in a fixed time interval. 

```{r}
load('Fingers.rda')
head(Fingers)
```

- 4 subjects (each subject is a block)
- 3 treatment groups (Placebo, Caffeine, Theobromine)
- Response: Finger Tapping Rate


--------------

#An experiment: One T-Test    

Question 1: Does caffeine speed up the tapping rate over the placebo?

\[H_0: \mu_{Caffeine} = \mu_{Placebo}\]
\[H_A: \mu_{Caffeine} \not= \mu_{Placebo}\]

Test statistic: $\bar{y}_{Caf} - \bar{y}_{Pla}$ or the standardized difference in  means

What assumptions need to hold for us to use the Student's T model?

```{r, tidy=TRUE}
require(mosaic)
mean(TapRate~Drug,data = Fingers)
39-22 #ybar1 - ybar2
t.test(Fingers$TapRate[Fingers$Drug == 'Caffeine'],Fingers$TapRate[Fingers$Drug == 'Placebo']) #standard t-test
```

**Main Issue:** These tapping rates in the groups are not independent! They are taken from the same individual!


If they were independent, then we would expect no pattern in the following plot.

```{r}
plot(Fingers$TapRate[Fingers$Drug == 'Caffeine'],Fingers$TapRate[Fingers$Drug == 'Placebo'])
```

The place where we used this assumption was in calculating the variance (and standard error), 

\[Var(\bar{y}_{Caf} - \bar{y}_{Pla}) = Var(\bar{y}_{Caf}) + Var(\bar{y}_{Pla}) \]

but now, if they aren't independent, then 

\[Var(\bar{y}_{Caf} - \bar{y}_{Pla}) = Var(\bar{y}_{Caf}) + Var(\bar{y}_{Pla}) - 2Cov(\bar{y}_{Caf},\bar{y}_{Pla}) \]

Rather than adjusting the variance, we could take differences within individuals,

```{r}
t.test(Fingers$TapRate[Fingers$Drug == 'Caffeine'],Fingers$TapRate[Fingers$Drug == 'Placebo'],paired=TRUE)
```

In this case, our data is **paired** so we can't just use a standard t-test due to the assumptions not holding. We used a **paired t-test**.

This is why it is VERY IMPORTANT TO CONSIDER THE ASSUMPTIONS of classical hypothesis tests!


--------------

#An experiment: Multiple T-Tests   

Question 2: Does caffeine speed up the tapping rate over the placebo? What about theobromine? How do the caffeine and theobromine compare?

Here we have three tests.

\[H_0: \mu_{Caffeine} = \mu_{Placebo}\]
\[H_0: \mu_{Theobromine} = \mu_{Placebo}\]
\[H_0: \mu_{Caffeine} = \mu_{Theobromine}\]

```{r}
t.test(Fingers$TapRate[Fingers$Drug == 'Caffeine'],Fingers$TapRate[Fingers$Drug == 'Placebo'],paired=TRUE)
```

```{r}
t.test(Fingers$TapRate[Fingers$Drug == 'Theobromine'],Fingers$TapRate[Fingers$Drug == 'Placebo'],paired=TRUE)
```

```{r}
t.test(Fingers$TapRate[Fingers$Drug == 'Caffeine'],Fingers$TapRate[Fingers$Drug == 'Theobromine'],paired=TRUE)
```

If we used a threshold of $\alpha = 0.05$ for all of these tests, what is the chance that I make at least one Type 1 error?

\[P(\text{at least one Type 1 error}) = 1 - P(\text{no Type 1 errors}) \]
\[= 1 - P(\text{no Type 1 error for one test})^3 \]
\[= 1 - (1-\alpha)^3 \]
\[= 1 - (1-0.05)^3 \]

```{r}
1 - (1-0.05)^3
```

If we were looking at 15 tests (comparing all pairs of 6 stimulants),

```{r}
1 - (1-0.05)^15
```

--------------

#Another data set: Multiple T-Tests  

```{r, include=FALSE,echo=FALSE}
library(mosaic)
teachers=read.delim("http://sites.williams.edu/rdeveaux/files/2014/09/teacher.txt")
```

Here are data from evaluations of 15 teachers on a scale from 1 to 20 by eight independent judges:

```{r}
boxplot(Score~Teacher,data=teachers,col=shuffle(rainbow(15)))
```

The standard t-tests show that there are some "significant" differences:

```{r}
options(width=130)
with(teachers,pairwise.t.test(Score,Teacher,p.adjust.method="none"))
sort(mean(Score~Teacher,data=teachers))
```

**So what would you conclude? Would you fire one or more of the teachers?**

**What possible errors are you making and what are the consequences of these decisions?**


**How do we control these possible errors?**

- One Approach
    - Adjust p-values
        - Bonferroni Correction
        - Tukey
        - Many others that try to control the "Family-wise Error Rate" (at least one Type 1 error) 

- Another Approach
     - Analysis of Variance F-test
        - $H_0: \mu_1 = \mu_2 = \mu_3 = \cdots = \mu_k$ with $\alpha$ as threshold.
        - In practice, just like fitting a multiple regression with dummy variables for groups.
        
```{r}
lm.1=lm(Score~Teacher,data = teachers)
summary(lm.1)
plot(residuals(lm.1)~predict(lm.1)) #check for equal variance and linearity
anova(lm.1)
summary(aov(lm.1))
```

Notice that the P-value is 0.323; this is not small enough to reject $H_0$. We don't have enough evidence for differences between teachers. 

With a P-value like this (after looking at boxplots to check assumptions), there’s no reason to try a modified pairwise set of t-tests. 

However, if this test does show evidence of some differences, then you are justified in running multiple comparisons (with appropriate P-value adjustment) on all the pairs.

*NOTE: The teacher data are not paired. If we were to do analysis of variance on the paired data, we would need to account for this lack of independence.*

--------------

#Another data set: Comparing Two Proportions

Let's look at our survey data. 

Is there a difference in the proportion of men and women that play a varsity sport??

```{r}
survey=read.delim("http://sites.williams.edu/rdeveaux/files/2014/09/Survey.13-15.txt")
```

```{r}
table(survey$Varsity,survey$Sex)
table(survey$Sex)
prop.table(table(survey$Varsity))
```

What would this table look like if $Sex$ and $Varsity$ were independent?

What numbers would you expect to see?


```{r,echo=FALSE}
chivs=chisq.test(table(survey$Varsity,survey$Sex),correct=FALSE)
chivs$expected
```
 
We'd expect the same overall proportion of No and Yes in both Female and Male columns (0.61, 0.38).


Let's compare the expected numbers with those that we observed in the table,

- Take difference between observed and expected
- Square the difference
- Divide each difference with the expected value (like standardizing it with respect to magnitudes)
- Add all of them together (X-squared value)


```{r}
chivs=chisq.test(table(survey$Varsity,survey$Sex),correct=FALSE)
sum((chivs$observed-chivs$expected)^2/chivs$expected)
```

Theoretical Sampling Distribution: Chi-squared distribution

```{r,echo=FALSE}
x = seq(0,20,by=.01)
plot(x,dchisq(x,df = 2),type='l',ylab='f(x)',xlab='x')
```

**What kind of properties does this sampling distribution model have? **

```{r}
chivs
```


--------------

#Another data set: Comparing More than Two Proportions

Suppose we have $k$ different groups/populations and we want to know if $p_1 = p_2 = ... = p_k$. 

This is called a **Test of Homogeneity of Groups**. For example, is the proportion of science majors the same at Swarthmore, Williams and Amherst? 

Let's look at some hypothetical data:

```{r results="asis",echo=FALSE}
library(xtable)
science=matrix(c(100,150,200,400,350,300,500,500,500),byrow=T,ncol=3)
colnames(science)=c("Swarthmore","Amherst","Williams")
rownames(science)=c("Science","Non-Science","Total")
science=data.frame(science)
print(xtable(science, digits=c(0,0,0,0)), type="html")
```

The test statistic is pretty logical. If the proportions were the same, we'd expect to see:

```{r results="asis",echo=FALSE}
expect=matrix(c(150,150,150,350,350,350,500,500,500),byrow=T,ncol=3)
colnames(expect)=c("Swarthmore","Amherst","Williams")
rownames(expect)=c("Science","Non-Science","Total")
expect=data.frame(expect)
print(xtable(expect, digits=c(0,0,0,0)), type="html")
```

The first table (without the totals) contains the **observed** values, $O_i$. The second one contains the **expected** values $E_i$. We take

$$ \sum{\frac{(O_i-E_i)^2}{E_i}}$$

If this is too big then the observed are too far away from the expected values.

This table has $2$ degrees of freedom. **Why?**

Compare to a $\chi^2_2$.

```{r}
chi=sum(((science[1:2,]-expect[1:2,])^2)/expect[1:2,])
chi
x = seq(0,20,by=.01)
plot(x,dchisq(x,df = 2),type='l',ylab='f(x)',xlab='x')
1-pchisq(chi,2)
```

We can have more than two categories in both the rows and columns.

```{r results="asis"}
socscience=matrix(c(100,150,200,150,200,100,250,150,200,500,500,500),byrow=T,ncol=3)
colnames(socscience)=c("Swarthmore","Amherst","Williams")
rownames(socscience)=c("Science","Arts","Social Science","Total")
socscience=data.frame(socscience)
print(xtable(socscience, digits=c(0,0,0,0)), type="html")
chisq.test(socscience)
```

Same test. Now the degrees of freedom are $(r-1) \times (c-1)$. 

-------------

##Another Test

When the rows and columns both signify representative values from two variables (measured on the same population), we call this a **Chi-square Test of Independence**, but the mechanics are the same as the Test for Homogeneity. 

Here's a test of independence on *Disease rate* and *Genotype*

```{r results="asis",echo=FALSE}
disease=matrix(c(268,199,42,807,759,184),ncol=2)
rownames(disease)=c("ins/ins","ins/del","del/del")
colnames(disease)=c("No Disease","Coronary Heart Disease")
disease=data.frame(disease)
print(xtable(disease, digits=c(0,0,0)), type="html")
```

```{r}
chisq.test(disease)
```

----------

##Last Test: Chi-square Goodness of fit

Finally, the Chi square test can be used in a different context -- to test whether a model fits by looking at its pmf. For example, are the digits of Pi uniformly distributed? 

Here are the frequencies of the digits 0-9 for the first 1,000,000 digits of pi. 

```{r}
digitsofpi=scan("http://sites.williams.edu/rdeveaux/files/2014/09/Digitsofpi.txt")
barplot(table(digitsofpi),col="lightblue")
table(digitsofpi)
sum((table(digitsofpi)-100000)^2/100000)
chisq.test(table(digitsofpi))
```

This is called the **Chi-square Goodness of Fit Test**.


