---
title: "Stat 201 - Lecture 1"
author: "Prof. Heggeseth"
date: "September 10, 2015"
output: slidy_presentation
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(dev='png',fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE,tidy=TRUE)
```

#Course Info
**More details in syllabus on [GLOW](glow.williams.edu)**

Prof. Brianna Heggeseth

bch2@williams.edu

Bronfman 287 (x5029)

##How to chat with me?

- Email (until 10pm)
- Scheduled Office Hours: Monday, Wednesday 2-3pm, Tuesday 3-4pm
- Schedule a meeting with me
- Invite me to coffee or dinner

#Who else can help?

- Your TA's: Bryan and Kelly
- De Veaux's TA's: Lia and Richard
- TA Sessions: Sun, Mon, Tues, Weds 8-10 pm Bronfman 119

#Assessments
##Reading Quizzes
- Online quizzes that you can complete **as many times as you would like** and the highest grade will be recorded. 
- You should work alone on these quizzes. 
- Goal: Become familiar with the basic vocabulary before class to get the most out of the time in lecture.

##Homework
- Homework will be assigned and submitted on GLOW.
-  Some problems (but not all) each week are selected (often at random) to be graded.
- Discuss and work on homework with your classmates, but you must submit your own write-up and obtain your own computer code and output.
- Complete all your homework in **RMarkdown**

#Should you do the homework?
Check out a **scatterplot** of homework and final exam grades for past courses.

```{r}
dat = read.csv('grades.csv')
plot(dat$Homework,dat$Final,pch=19)
```

#Should you do the homework?
Check out a **scatterplot** of homework and final exam grades for past courses.

```{r }
plot(dat$Homework,dat$Final,xlim=c(0.8,1),ylim=c(.6,1),pch=19)
```

#Assessments
##Exams
- Two midterms (in-class and take-home portion for each)
- Final exam (in-class cumulative)

Exams will make up 80% of your grade.


**Read the rest of the syllabus online for more infomation**

#You are Hired
##Thought Experiment
The City of San Francisco hires you as a consultant to provide concrete guidelines to help them use their resources more effectively and efficiently to reduce crime.


##Data
You have access to all reported crimes in the city of SF since 2003 (available at: https://data.sfgov.org/Public-Safety/SFPD-Incidents-from-1-January-2003/tmnf-yvry).

- What are the **cases** (what do the rows represent)?
- What **variables** do we have (what characteristics do we have on each case)?
- Are the variables **categorical** (values that are names of categories) or **quantitative** (measured numerical values with units)?


#First Step 

- With someone next to you, come up with 5 specific research questions that you want to address with the data. 
- Share your ideas with the class!

#Explore Data for 2014

Typically the most interesting (and most relevant) questions involve more than one variable, but let's start with one variable at a time.

Go to: http://galileo.williams.edu:3838/users/bch2/SFCrimeEDA (This is a Shiny App made in RStudio, the software that we will be using; all graphs made using R)

Be patient as it loads (it is a large data set).

- Click through the different variables; there are appropriate graphics and numerical for that type of variable (categorical or quantitative). **What did you learn?**

- Now click on Bivariate Exploration. Explore the different combinations of variables. **What did you learn?**

#Explore Data for 2014

Go to: http://galileo.williams.edu:3838/users/bch2/SFCrimeMap

- Click through Points, Neighborhoods, and Heat Map. **Are there more questions that you have thought of?**
- You can filter the data and look at subsets (by day of the week, time of day, Crime Type, etc.). 

**Over the weekend, explore this data set and think about guidelines for the SF Police Department. We'll come back to this data set.**

#Key Concepts

**Interesting relationships involve more than one or even two variables.**

First, we will talk about summarizing 1-2 variables. 

#Categorical Variables
##One Variable (Univariate)
The main graphical summary for one categorical variable is a **bar plot.** 

- The **height** of the bars is the only part that shows the data (width is meaningless here). 
- The height can either represent the **frequency** (count of cases) or the **relative frequency** (proportion of cases).

```{r}
dat <- read.csv('SFPD.csv')
(tab <- sort(table(dat$Category), decreasing=TRUE)) #Frequency table
barplot(tab[1:10], las = 2, cex.names= .5, col='darkblue', main='Bar Plot (Frequency)',ylab='Frequency')
barplot(prop.table(tab)[1:10], las = 2, cex.names= .5, col='darkblue', main='Bar Plot (Relative Frequency)',ylab='Relative Frequency')
```

A good summary of why **I hate pie charts**: http://www.businessinsider.com/pie-charts-are-the-worst-2013-6 (make sure to read if you like pie charts!)

#Categorical Variables
##Two Variables (Bivariate)
The best (my opinion) graphical summary for the relationship between two categorical variables is a **mosaic plot**.

- The **relative height** of the bars shows the conditional distribution (relative frequency within subsets).
- The **width** of the bars shows the marginal distribution (relative frequency of the X variable, ignoring the other variable).

```{r, tidy=TRUE}
dat$DayOfWeek <- factor(dat$DayOfWeek,levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
(tab2 <- table(dat$DayOfWeek,dat$PdDistrict)) #Contingency table
mosaicplot(tab2, las = 2, col=rainbow(9), main='Mosaic Plot (Condition on Day)')
(tab3 <- table(dat$PdDistrict,dat$DayOfWeek)) #Contingency table
mosaicplot(tab3, las = 2, col=rainbow(7), main='Mosaic Plot (Condition on District)')
```



#Quantitative Variables
##One Variable (Univariate)
The main graphical summary for one quantitative variable is a **histogram**. 

- The **height** of the bars shows either the **frequency within intervals** or the **density** (count per width of interval)
- The x-axis is a number line and bars do not have to be of equal width (R defaults to equal width, but you can change how many intervals and how large they are)

```{r, tidy=TRUE}
bodyfat <- read.delim("http://sites.williams.edu/rdeveaux/files/2014/09/bodyfat.txt")
hist(bodyfat$Wrist, col="lightblue",main="",ylab="# of men",xlab='Wrist circumference (cm)')

hist(bodyfat$Wrist, col="lightblue",main="",ylab="# of men",xlab='Wrist circumference (cm)', breaks = seq(15,22,by = 0.05)) #Gaps are meaningful
```

We focus on three aspects of the distribution:

- Shape: is it **symmetric**? or **skewed to the right or left** (skewed to the **tail**)? How many **modes** do you see?
- Center: where is a typical value located?
- Spread (or variation): how spread out is the distribution?

Also, we ask
– are there any unusual features such as outliers or gaps?

```{r, tidy=TRUE}
ceo <- read.delim("http://sites.williams.edu/rdeveaux/files/2014/09/CEO_Salary_2012.txt")
hist(ceo$X1.Year.Pay...mil.,col="light green",main="",xlab="Salary (in $M)",ylab="# of CEOs")
```

For center we have some choices for numerically summarizing it:

- **Mean**: the sum of the observations divided by the sample size, $\bar{y} = \frac{\sum^n_{i=1}y_i}{n}$
    + Sensitive to outliers, but efficiently uses all the data
- **Median**: the "middle" value. The number for which half the values are below and half are above.
    + Insensitive to outliers, but doesn’t use all the actual values
- **Trimmed means**:  Drop the lowest and highest k% and take the mean of the rest
    + A good compromise, but not widely used.

```{r, tidy=TRUE}
summary(bodyfat$Wrist)
mean(bodyfat$Wrist)
median(bodyfat$Wrist)
mean(bodyfat$Wrist,trim = 0.05) #Trim 5% from both ends


summary(ceo$X1.Year.Pay...mil.)
mean(ceo$X1.Year.Pay...mil.)
median(ceo$X1.Year.Pay...mil.)
mean(ceo$X1.Year.Pay...mil.,trim = 0.05) #Trim 5% from both ends
```

For spread we have several choices to numerically describe it:

- **Range**: the max - the min.
    + Sensitive to the outliers since it’s the difference of the extremes
- **IQR**: the interquartile range : Q3 - Q1 (75th percentile - 25th percentile).
    + Like the median. Less sensitive, but less "efficient"
- **Standard deviation**: Root mean squared deviations from mean, $s = \sqrt{\frac{\sum^n_{i=1}(y_i-\bar{y})^2}{n-1}}$
    + Uses all the data but very sensitive to outliers and skewed data.

```{r, tidy=TRUE}
diff(range(bodyfat$Wrist)) #range gives max and min; take difference
IQR(bodyfat$Wrist)
sd(bodyfat$Wrist)

diff(range(ceo$X1.Year.Pay...mil.))
IQR(ceo$X1.Year.Pay...mil.)
sd(ceo$X1.Year.Pay...mil.) 
```

#Summary for Quantitative Variables

For a quantitative variable,

- Use a histogram to display the distribution and describe the shape and any unusual features.
- For “well behaved” distributions (symmetric, unimodal, no outliers) use the mean and standard deviation to describe the center and spread.
- For others, use the IQR and median. You can report both mean and median, but it’s usually a good idea to state why.

