---
title: "Stat 201 - Lecture 9"
author: "Prof. Heggeseth"
date: "October 8, 2015"
output: slidy_presentation
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(dev='png',fig.path='Figs/',echo=TRUE,message=FALSE, warning=FALSE,message=FALSE,tidy=TRUE,fig.width = 6,fig.height=6)
```

----------

#Which Variables to Include?

We saw last lecture how including more variables can help us get a more "accurate" understanding of relationships because we are keeping variables constant ("comparing apples to apples").

Let's throw all the body measurements in to see the relationships with Percent Body Fat.

```{r}
library(car)
bodyfat = read.delim("http://sites.williams.edu/rdeveaux/files/2014/09/bodyfat.txt")

bodyfat$BMI = (bodyfat$Weight*0.453592) /(bodyfat$Height*0.0254)^2

lm.full = lm(Pct.BF~.,data=bodyfat)
summary(lm.full)
avPlots(lm.full)
```

How do we choose which variables are "important"?

----------

#Criteria

- Can't have more variables than there are cases ($n<p$).
- Want small SSE (better predictions).
- Want $R^2$ high (Issue: $R^2$ will always increase or stay the same with more variables)
- Want adjusted $R^2$ high (we add a little penalty for having too many variables in the model)
- Want variables included to "be associated with" body fat even after accounting for other variables (they need to contribute!)


----------

#Backwards Elimination

Starting with a full model, we could eliminate variables one at a time that don't "help" the model. 

Run models leaving out one variable at a time, and then compare by a criterion.


Criterion Options:

- Use adjusted $R^2$
- t-values (SD's away from 0 to account for sampling variablilty)
- AIC, which is SSE with a penalty for too many variables (want small)


```{r}
lm.step=step(lm.full) #Default: uses AIC
summary(lm.step)
```


----------

#Review

##Graphical and Numerical Summaries

**One Categorical Var**

- Frequency Tables (and Relative Frequencies)
- Barplots

**One Quantitative Var**

- 5-number summary
- Center: median, mean, trimmed mean
- Spread: IQR, sd, range
- Z-scores
- How usual are z-scores? (Chebyshev and 68-95-99.7 rule)
- Shifting and Scaling impact on summary statistics
- Histograms (shape, QQ plots to compare to Normal)
- Boxplots (outliers..)

**Two Categorical Var**

- Contingency table
- Mosaic plots

**One Categorical Var, One Quantitative Var**

- Mean, SD, Median, IQR within groups
- Multiple Boxplots

**Two Quantitative Var**

- Scatterplots
- Correlation
- Shifting and Scaling impact on correlation
- Regression Lines (simple)

##Linear Models

**Regression**

- Least Squares (minimizing SSE)
- Residuals
- Interpreting Slope and Intercept
- Standard deviation of residuals (idea of prediction error)
- Regressing to the mean
- Impact of outliers
- Conditions and Diagnostics (residual plots, added variable plots)
- Transformations (Ladder of Powers and Box-Cox)

##Sampling Variability

- Sampling Distribution from Population (center and spread)
- Bootstrapping Sampling Distribution (center and spread)
- Bootstrap Confidence Intervals
- "Shuffling" groups to test a hypothesis (assume no difference)


